Late Pleistocene stone artefact technology at Mau A, Yen Bai, Northern Vietnam

### 2022-03-05 analysis/paper/paper.Rmd
improvements to plots, many minor edits to the text, remove unused code and outdated comments

---
title: "Late Pleistocene stone artefact technology at Mau A, Yen Bai, Norther Vietnam"
author:
  - Quinn Habedank:
      email: fl@oneoeurhg.edu
      institute: [UofO]
      correspondence: true
  - Ben Marwick:
      email: fl@another.edu
      institute: [UofA]
      correspondence: false
institute:
  - UofO: University of One Place
  - UofA: University of Another Place
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

<!-- BM: In this section we need to review previous work on investigating flake shape and identifying what affects flake shape, not just in Southeast Asia, but anywhere in the world -->

@@

<!-- QH: Trying to think of a way to make this figure work with the current figure numbering system we have in bookdown. Adding in a number that starts before 1.1 for now -->
## Methods

![Figure 1.0. Schematic figure showing both the intial measurements that were used to generate our landmarks and the location of our landmarks on a hypothetical flake. A, B, C, D, and E are the top horizontal measurement, width at a quarter of the maximum length, maximum dimension length, width at a quarter of the maximum length, and width at three quarter of the maximum length respectively. Points 1 through 9 are all of the landmarks generated from measurements A through E.  ](./test_graphic.png)

@@

## PCA Analysis on Dorsal Cortex Groups

@@

# TODO: find a meaningful way to break this up into groups
# pass these groups onto the PCA for drawing ellipses

# this block of code below shows one way to discretize cortex percentage
# into three groups. We need to explore other cutoffs, and other grouping
# variables besides dorsal cortex. This code below is a good template for
# starting these explorations
```

```{r, fig.cap="PCA biplot for flake shape and dorsal cortex class."}

ma_flakes_tr_with_groups <-
  ma_flakes_tr %>%
  slice(-c(1152,  606, 1147, 1125,  756 ,1023,  457,
                 813,  415,  877,  677, 1055,  239,  811,  601,  861,
                  794,  103 , 598 ,1110 , 906 ,  59 ,   4)) %>% 
 # filter(!is.na(dorsal_cortex_perc)) %>% 
  mutate(dorsal_cortext_group =
           case_when(
             between(dorsal_cortex_perc, 0, 0) ~ "Tertiary",
             between(dorsal_cortex_perc, 1, 50) ~ "Secondary",
             between(dorsal_cortex_perc, 51, 100) ~ "Primary"
           ))

@@

ma_flakes_tr_with_groups <-
  ma_flakes_tr_with_groups %>% 
  mutate(mass_group = as.factor(test_result_mass$cluster))

# ggplot2-based plot of the PCA results using a factoextra function
cortex_biplot <- fviz_pca_ind(pca.geo,
             geom.ind = "point", # show points without labels
            #  line below sets the grouping column for colouring the points
             habillage = ma_flakes_tr_with_groups$dorsal_cortext_group,
             palette = "npg", #color palette, adjust to preferences
             addEllipses = T, #adds ellipses or convex hulls
             ellipse.type = "convex", #convex hulls
             mean.point = F, #don’t show average point
             legend.title = "Dorsal Cortex Class") +  #adjust accordingly)
             coord_fixed(ratio = 1)+ #adjustable for variation-weighted plots
  theme(plot.title = element_blank())

@@

```{r, fig.cap="PCA biplot for flake shape and excavation unit."}
# Excavation unit PCA
# ggplot2-based plot of the PCA results using a factoextra function
excavation_unit_biplot <- fviz_pca_ind(pca.geo,
             geom.ind = "point", # show points without labels
            #  line below sets the grouping column for colouring the points
             habillage = as.factor(ma_flakes_tr_with_groups$Unit),
             palette = viridis::cividis(13), # color palette, adjust to preferences
             addEllipses = T, #adds ellipses or convex hulls
             ellipse.type = "convex", #convex hulls
             mean.point = F, #don’t show average point
             legend.title = "Excavation\nunit",
             title = "Change over time in flake shape") +  #adjust accordingly)
             coord_fixed(ratio = 1)  + #adjustable for variation-weighted plots
             theme(panel.background = element_rect(fill='white',
                                        colour = "white"),
              plot.background = element_rect(fill='white',
                                       colour = "white")) +
    theme(plot.title = element_blank())
```

<!-- This portion is lifted largely from the ISKM presentation. Might well reword it later, but it will do for now -->

```{r, fig.cap="PCA biplot for flake shape and mass cluster groups"}

# Mass PCA
# ggplot2-based plot of the PCA results using a factoextra function
mass_biplot <- fviz_pca_ind(pca.geo,
             geom.ind = "point", # show points without labels
            #  line below sets the grouping column for colouring the points
             habillage = ma_flakes_tr_with_groups$mass_group,
             palette = viridis::viridis(6), #color palette, adjust to preferences
             addEllipses = T, #adds ellipses or convex hulls
             ellipse.type = "convex", #convex hulls
             mean.point = F, #don’t show average point
             legend.title = "Mass\ncluster",  #adjust accordingly
            title = "Flake mass and flake shape") + 
             coord_fixed(ratio = 1) + #adjustable for variation-weighted plots
             theme(panel.background = element_rect(fill='white',
                                        colour = "white"),
              plot.background = element_rect(fill='white',
                                       colour = "white")) +
    theme(plot.title = element_blank())
```

<!-- Below are my attempts to create a nice looking plot_grid of the three biplots. I was not able to get it to display properly, so I am just printing off each biplot individually all in a row as a temporary measure.  -->
```{r}
library(cowplot)

# plot_grid(
#  cortex_biplot, NULL, NULL,  excavation_unit_biplot,  mass_biplot, NULL,
#  ncol = 2,
#  labels = c("A", NULL, NULL,"B","C",NULL),
#  label_size = 12,
#  align = "v",
#  rel_heights = c(1.1,
#                  1,
#                  1.1,
#                  4,
#                  1,
#                  1),
#  rel_widths = c(2,
#                 2,
#                 2,
#                 4,
#                 2,
#                 1))

#cortex_biplot + theme(plot.margin = unit(c(3,3,3,3), "lines"))
#excavation_unit_biplot
#mass_biplot

#plot_grid(
#  plot_a, NULL, plot_b
#  , nrow = 1
#  , rel_widths = c(1, 0.05, 1)
#  , labels = c("A", "", "B")
#)

    #      axis = "none",
   #       scale = 1,
   #       nrow = 3,
    #      ncol = 1,
   #       rel_widths = c(1,1,1),
   #       rel_heights = c(1,1,1)

@@

```{r cortex-biplot, fig.cap="PCA biplot for flake shape and dorsal cortex class."}
cortex_biplot
```

```{r excavation-unit-biplot, fig.cap="PCA biplot for flake shape and excavation unit."}
excavation_unit_biplot
```

```{r mass-biplot, fig.cap="PCA biplot for flake shape and mass cluster groups"}
mass_biplot
```

Figure \@ref(fig:cortex-biplot) shows flake shape (as defined within the first two principal components) in relation to dorsal cortex class. The primary class of dorsal cortex, representing initial, unretouched flakes in the reduction sequence, has the smallest range of shape variation. Secondary flakes, which have been retouched and thus have less dorsal cortex than the primary flakes, show a slightly larger range of shape variation. Lastly, the tertiary flakes at the end of the reduction sequence show the greatest amount of shape variation.

@@

Figure \@ref(fig:excavation-unit-biplot) shows the flake shapes projected on the first two principal components, and colored according to the excavation unit that they were recovered from. Excavation unit one is near the surface, and unit 13 is near the base. The general picture here is that shape variation changes little through the deposit. Units with fewer flakes, such as 13, show lower shape variation.

In Figure \@ref(fig:mass-biplot) we see that greater shape variation is found in the lightest mass classes. The two lightest mass clusters, one and two, are in purple and blue respectively. These first two mass classes span the largest area on the biplot, signifying a larger degree of variation overall.

```{r contribution-plots, fig.cap="Plots showing the contribution of landmarks to the overall shape variation among flakes."}
# let's look at all the PCs and how much each original # variable contributes to them

#view how much variation is described by each PC (ordered from PC1 to last PC)
# enframe(100*pca.geo$sdev^2/sum(pca.geo$sdev^2))

#get meaningful PCs
# getMeaningfulPCs(pca.geo$sdev^2, n = length(lands_no_outliers_names))

#estimate contribution (%) of landmarks on PC1
PC1_contrib <- fviz_contrib(pca.geo, choice = "var", axes = 1)
# take a look at the plot
# PC1_contrib

LM_contrib_PC1 <- data.frame(LM = rep(NA, nrow(PC1_contrib$data)/2),
                             PC1_contrib = rep(NA, nrow(PC1_contrib$data)/2))
for (i in 1:(nrow(PC1_contrib$data)/2)) {
  LM_contrib_PC1$PC1_contrib[i] <- sum(PC1_contrib$data[(i*2-1):(i*2),]$contrib)
  LM_contrib_PC1$LM[i] <- as.character(i)
}

# estimate contribution (%) of landmarks on PC2
PC2_contrib <- fviz_contrib(pca.geo, choice = "var", axes = 2)
# take a look at the plot
# PC2_contrib

LM_contrib_PC2 <- data.frame(LM = rep(NA, nrow(PC2_contrib$data)/2),
                             PC2_contrib = rep(NA, nrow(PC2_contrib$data)/2))
for (i in 1:(nrow(PC2_contrib$data)/2)) {
  LM_contrib_PC2$PC2_contrib[i] <- sum(PC2_contrib$data[(i*2-1):(i*2),]$contrib)
  LM_contrib_PC2$LM[i] <- as.character(i)
}

# examine landmarks with major contribution (%) to PC1 and PC2
# important_LMs_PC1 <- LM_contrib_PC1[LM_contrib_PC1$PC1_contrib > 100/(nrow(PC1_contrib$data)/2),]
# important_LMs_PC1[order(decreasing = T, important_LMs_PC1$PC1_contrib),]
# important_LMs_PC2 <- LM_contrib_PC2[LM_contrib_PC2$PC2_contrib > 100/(nrow(PC2_contrib$data)/2),]
# here is a table that summarises the most important landmarks in the PCA
# important_LMs_PC2[order(decreasing = T, important_LMs_PC2$PC2_contrib),]

@@

plot_grid(PC1_contrib +
          scale_x_discrete(labels = 
                           as.factor(c("X Value: Top Left",
                          "Y Value: Top Left",
                          "X Value: Mid-Upper Left",
                          "Y Value: Mid-Upper Left",
                          "X Value: Mid-Lower Left",
                          "Y Value: Mid-Lower Left",
                          "X Value: Bottom Left",
                          "Y Value: Bottom Left",
                          "X Value: Distal Tip",
                          "Y Value: Distal Tip",
                          "X Value: Bottom Right",
                          "Y Value: Bottom Right",
                          "X Value: Mid-Lower Right",
                          "Y Value: Mid-Lower Right",
                          "X Value: Mid-Upper Right",
                          "Y Value: Mid-Upper Right",
                          "X Value: Top Right",
                          "Y Value: Top Right")))+
             theme(axis.text = element_text(size = 5)),
          PC2_contrib + 
          scale_x_discrete(labels = 
                           as.factor(c("X Value: Top Left",
                           "Y Value: Top Left",
                           "X Value: Mid-Upper Left",
                           "Y Value: Mid-Upper Left",
                           "X Value: Mid-Lower Left",
                           "Y Value: Mid-Lower Left",
                           "X Value: Bottom Left",
                           "Y Value: Bottom Left",
                           "X Value: Distal Tip",
                           "Y Value: Distal Tip",
                           "X Value: Bottom Right",
                           "Y Value: Bottom Right",
                           "X Value: Mid-Lower Right",
                           "Y Value: Mid-Lower Right",
                           "X Value: Mid-Upper Right",
                           "Y Value: Mid-Upper Right",
                           "X Value: Top Right",
                           "Y Value: Top Right")))+
             theme(axis.text = element_text(size = 5))
  )

@@

Figure \@ref(fig:contribution-plots) shows the relative contribution of each of the original 18 variables put through our PCA. Looking at the supplementary table, we can see that the first principal component is primarily composed of the x-value for the distal tip landmark, with smaller contributions from the y-values from the two upper-most and mid-upper points. The second primary dimension is composed largely of the y-values for bottom-most and upper-most points.

# Discussion
  <!-- This portion partially consists of material from the ISKM presentation. Might want to reword things even more. Moreover, this section needs citations, which will bolster the overall content of this section. -->

  Our data visualizations reveal some noteworthy trends regarding the degree of core reduction. Figure \@ref(fig:excavation-unit-biplot) shows very little variation over time, with lots of overlap between groups in the biplot. Units with fewer flakes, such as 13, show lower shape variation. This generally confirms our conclusion from the radiocarbon ages that the deposit accumulated relatively quickly, and the assemblage represents knapping events occurring at a similar time with little, if any perceptible change over time. By contrast, a more definitive difference between groups in seen in Figure \@ref(fig:mass-biplot), where there is a high degree of shape variation in the lightest flakes. This reflects the production of different types and shapes of cores as the flakes become smaller and thus lighter. This is further supported by Figure \@ref(fig:cortex-biplot), where tertiary flakes with little dorsal cortex show the highest shape variation. This supports our interpretation of the previous result, with flakes from later in the reduction process showing greater shape variation as different types of cores were produced at the site.

@@

Looking at \@ref(fig:contribution-plots), the landmarks which contributed the most to each principal component were all located on the left side of our artefacts. This means that linking our initial flake measurements (maximum length, top width etc.) to either of the first two principal components is nearly impossible, as every one of our initial flake measurements (with the)

one can see that there is no consistent pattern in which certain initial flake measurements  align closely with either the first or second principal components. Interestingly, the landmarks which contributed the most to each principal component were all located on the left side of our artefacts..

The lack of distinct morphological groupings and the increased variability of shape in the smaller, tertiary flakes is consistent with previous studies. Other studies which investigated the morphology of flake outlines from South East Asia concluded that there is no significant relationship between flake outline and flake use (Borel et al., 2013; Borel et al., 2017). While our analysis not rely upon outlines, both outlines and landmarks are ultimately methods for measuring the morphology of flaked stone tools, and as such, the same models which explain outline variability can be cautiously applied to our analysis. The large amount of variability in flake shape, especially in the tertiary flakes, suggests that the knappers were not looking to create a specific shape of flake, with the amount of variability increasing as flakes are struck off of increasingly differed cores. This behavior is consistent with processes of expedient knapping, where a large number of flakes are created in a relatively free-form manner and useful flakes are picked up and used after knapping is finished (Holdaway & Douglass, 2012; McCall, 2012). Processes of expedient knapping suggest that stone tools were of secondary importance in the material culture of the knappers (Sillitoe & Hardy, 2003). Elsewhere it has been suggested stone tools were primarily used to create bamboo tools in South East Asia (Mijares, 2001). A reliance upon bamboo tools, with stone flakes merely being an intermediary to produce bamboo tools, would explain the great variability in the lithic assemblages at Mau A.

>>>>

---
title: "Late Pleistocene stone artefact technology at Mau A, Yen Bai, Northern Vietnam"
author:
  - Quinn Habedank:
      email: fl@oneoeurhg.edu
      institute: [UofO]
      correspondence: true
  - Ben Marwick:
      email: fl@another.edu
      institute: [UofA]
      correspondence: false
institute:
  - UofO: University of One Place
  - UofA: University of Another Place
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      number_sections: false
      global_numbering: true
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

<!-- BM: In this section we need to review in very precise detail, previous work on investigating flake shape and identifying what affects flake shape, not just in Southeast Asia, but anywhere in the world -->

@@

## Methods

```{r, schematic-figure, fig.cap="Schematic figure showing both the intial measurements that were used to generate our landmarks and the location of our landmarks on a hypothetical flake. A, B, C, D, and E are the top horizontal measurement, width at a quarter of the maximum length, maximum dimension length, width at a quarter of the maximum length, and width at three quarter of the maximum length respectively. Points 1 through 9 are all of the landmarks generated from measurements A through E."}
# knitr::include_graphics(here::here("analysis/flake-measurements-schematic-figure.png"))
```

@@

## Principal Components Analyses exploring flake shape variation

@@

```{r, mass-dorsal-grouping-and-cluster-analysis}

ma_flakes_tr_with_groups <-
  ma_flakes_tr %>%
  slice(-c(1152,  606, 1147, 1125,  756 ,1023,  457,
                 813,  415,  877,  677, 1055,  239,  811,  601,  861,
                  794,  103 , 598 ,1110 , 906 ,  59 ,   4)) %>% 
  mutate(dorsal_cortext_group =
           case_when(
             between(dorsal_cortex_perc, 0, 0) ~ "Tertiary",
             between(dorsal_cortex_perc, 1, 50) ~ "Secondary",
             between(dorsal_cortex_perc, 51, 100) ~ "Primary"
           ))

@@

ma_flakes_tr_with_groups <-
  ma_flakes_tr_with_groups %>% 
  mutate(mass_group = as.factor(test_result_mass$cluster))
```

```{r pca-cortex-groups}

# to adjust the legend size for all three plots
legend_key <- 0.25
legend_title_size <- 6
legend_text_size <- 4

# ggplot2-based plot of the PCA results using a factoextra function
cortex_biplot <- fviz_pca_ind(pca.geo,
             geom.ind = "point", # show points without labels
            #  line below sets the grouping column for colouring the points
             habillage = ma_flakes_tr_with_groups$dorsal_cortext_group,
             palette = "npg", #color palette, adjust to preferences
             addEllipses = T, #adds ellipses or convex hulls
             ellipse.type = "convex", #convex hulls
             mean.point = F, #don’t show average point
             legend.title = "Dorsal\ncortex\nclass") +  #adjust accordingly)
             coord_fixed(ratio = 1) + #adjustable for variation-weighted plots
  theme(plot.title = element_blank(),
        axis.text = element_text(size = legend_text_size),
        axis.title = element_text(size = legend_title_size),
        legend.key.size = unit(legend_key, 'cm'), #change legend key size
        legend.key.height = unit(legend_key, 'cm'), #change legend key height
        legend.key.width = unit(legend_key, 'cm'), #change legend key width
        legend.title = element_text(size = legend_title_size), #change legend title font size
        legend.text = element_text(size = legend_text_size))

@@

```{r, pca-excavation-unit}
# Excavation unit PCA
# ggplot2-based plot of the PCA results using a factoextra function
excavation_unit_biplot <- fviz_pca_ind(pca.geo,
             geom.ind = "point", # show points without labels
            #  line below sets the grouping column for colouring the points
             habillage = as.factor(ma_flakes_tr_with_groups$Unit),
             palette = viridis::cividis(13), # color palette, adjust to preferences
             addEllipses = T, #adds ellipses or convex hulls
             ellipse.type = "convex", #convex hulls
             mean.point = F, #don’t show average point
             legend.title = "Excavation\nunit",
             title = "Change over time in flake shape") +  #adjust accordingly)
             coord_fixed(ratio = 1)  + #adjustable for variation-weighted plots
             theme(panel.background = element_rect(fill='white',
                                        colour = "white"),
              plot.background = element_rect(fill='white',
                                       colour = "white")) +
    theme(plot.title = element_blank(),
        axis.text = element_text(size = legend_text_size),
        axis.title = element_text(size = legend_title_size),
        legend.key.size = unit(legend_key, 'cm'), #change legend key size
        legend.key.height = unit(legend_key, 'cm'), #change legend key height
        legend.key.width = unit(legend_key, 'cm'), #change legend key width
        legend.title = element_text(size = legend_title_size), #change legend title font size
        legend.text = element_text(size = legend_text_size)) 
```

```{r, pca-mass-cluster-groups}

# Mass PCA
# ggplot2-based plot of the PCA results using a factoextra function
mass_biplot <- fviz_pca_ind(pca.geo,
             geom.ind = "point", # show points without labels
            #  line below sets the grouping column for colouring the points
             habillage = ma_flakes_tr_with_groups$mass_group,
             palette = viridis::viridis(6), #color palette, adjust to preferences
             addEllipses = T, #adds ellipses or convex hulls
             ellipse.type = "convex", #convex hulls
             mean.point = F, #don’t show average point
             legend.title = "Mass\ncluster",  #adjust accordingly
            title = "Flake mass and flake shape") + 
             coord_fixed(ratio = 1) + #adjustable for variation-weighted plots
             theme(panel.background = element_rect(fill='white',
                                        colour = "white"),
              plot.background = element_rect(fill='white',
                                       colour = "white")) +
      theme(plot.title = element_blank(),
        axis.text = element_text(size = legend_text_size),
        axis.title = element_text(size = legend_title_size),
        legend.key.size = unit(legend_key, 'cm'), #change legend key size
        legend.key.height = unit(legend_key, 'cm'), #change legend key height
        legend.key.width = unit(legend_key, 'cm'), #change legend key width
        legend.title = element_text(size = legend_title_size), #change legend title font size
        legend.text = element_text(size = legend_text_size)) 
```

```{r three-pca-biplots, fig.cap="PCA biplots showing (A) variation in flake shape with dorsal cortex class, (B) depth below surface, and (C) mass cluster", fig.height=6}
library(cowplot)

plot_grid(
 cortex_biplot,            # A
 excavation_unit_biplot,   # B
 mass_biplot,              # C
 ncol = 1,
 align = "v", 
 axis = "lr",
 labels = "AUTO")

@@

Figure \@ref(fig:three-pca-biplots)A shows flake shape (as defined within the first two principal components) in relation to dorsal cortex class. The primary class of dorsal cortex, representing initial, unretouched flakes in the reduction sequence, has the smallest range of shape variation. Secondary flakes, which have been retouched and thus have less dorsal cortex than the primary flakes, show a slightly larger range of shape variation. Lastly, the tertiary flakes at the end of the reduction sequence show the greatest amount of shape variation.

@@

Figure \@ref(fig:three-pca-biplots)B shows the flake shapes projected on the first two principal components, and colored according to the excavation unit that they were recovered from. Excavation unit one is near the surface, and unit 13 is near the base. The general picture here is that shape variation changes little through the deposit. Units with fewer flakes, such as 13, show lower shape variation.

In Figure \@ref(fig:three-pca-biplots)C we see that greater shape variation is found in the lightest mass classes. The two lightest mass clusters, one and two, are in purple and blue respectively. These first two mass classes span the largest area on the biplot, signifying a larger degree of variation overall.

```{r contribution-plots, fig.cap="Plots showing the contribution of landmarks to the first two Principal Components (A: PC1, B: PC2) that summarise the overall shape variation among flakes. The dashed red horizontal line corresponds to the expected value if the contribution from each landmark was uniform.", fig.height = 2}

# let's look at all the PCs and how much each original # variable contributes to them

# estimate contribution (%) of landmarks on PC1
PC1_contrib <- fviz_contrib(pca.geo, 
                            choice = "var", 
                            axes = 1) +
   theme(plot.title = element_blank(),
        axis.text = element_text(size = legend_text_size),
        axis.title = element_text(size = legend_title_size),
        legend.key.size = unit(legend_key, 'cm'), #change legend key size
        legend.key.height = unit(legend_key, 'cm'), #change legend key height
        legend.key.width = unit(legend_key, 'cm'), #change legend key width
        legend.title = element_text(size = legend_title_size), #change legend title font size
        legend.text = element_text(size = legend_text_size))

# estimate contribution (%) of landmarks on PC2
PC2_contrib <- fviz_contrib(pca.geo, 
                            choice = "var", 
                            axes = 2) +
  theme(plot.title = element_blank(),
        axis.text = element_text(size = legend_text_size),
        axis.title = element_text(size = legend_title_size),
        legend.key.size = unit(legend_key, 'cm'), #change legend key size
        legend.key.height = unit(legend_key, 'cm'), #change legend key height
        legend.key.width = unit(legend_key, 'cm'), #change legend key width
        legend.title = element_text(size = legend_title_size), #change legend title font size
        legend.text = element_text(size = legend_text_size))

# this table will allow us to update the plot axis tick mark labels with human-readable
# labels
long_labels <- 
  tibble::tribble(       ~long_labels,             ~short_labels, 
                 "X Value: Top Left",         "1.X",
                 "Y Value: Top Left",         "2.X",
                 "X Value: Mid-Upper Left",    "...",
                 "Y Value: Mid-Upper Left",    "...", 
                 "X Value: Mid-Lower Left",    "...", 
                 "Y Value: Mid-Lower Left",    "...", 
                 "X Value: Bottom Left",       "...",
                 "Y Value: Bottom Left",      "...", 
                 "X Value: Distal Tip",       "...",  
                 "Y Value: Distal Tip",       "...", 
                 "X Value: Bottom Right",     "...", 
                 "Y Value: Bottom Right",     "...", 
                 "X Value: Mid-Lower Right",   "...", 
                 "Y Value: Mid-Lower Right",    "...", 
                 "X Value: Mid-Upper Right",    "...", 
                 "Y Value: Mid-Upper Right",    "...", 
                 "X Value: Top Right",         "...",  
                 "Y Value: Top Right",         "..."   
                 )

@@

plot_grid(PC1_contrib +
             theme(axis.text = element_text(size = 5)),
          PC2_contrib + 
             theme(axis.text = element_text(size = 5)),
          labels = "AUTO"
  )

@@

Figure \@ref(fig:contribution-plots) shows the relative contribution of each of the original 18 variables put through our PCA. The first principal component is primarily composed of the x-value for the distal tip landmark, with smaller contributions from the y-values from the two upper-most and mid-upper points. The second principal component is composed largely of the y-values for bottom-most and upper-most points.

## Principal Components Analyses exploring core shape variation

# Discussion

Flake shape variation generally shows limited correspondence with other factors. Figure \@ref(fig:excavation-unit-biplot)B shows very little variation over time, with lots of overlap between groups in the biplot. Units with fewer flakes, such as 13, show lower shape variation. This generally confirms our conclusion from the radiocarbon ages that the deposit accumulated relatively quickly, and the assemblage represents knapping events occurring at a similar time with little, if any perceptible change over time. By contrast, a more definitive difference between groups in seen in Figure \@ref(fig:excavation-unit-biplot)C, where there is a high degree of shape variation in the lightest flakes. This reflects the production of different types and shapes of cores as the flakes become smaller and thus lighter. This is further supported by Figure \@ref(fig:excavation-unit-biplot)A where tertiary flakes with little dorsal cortex show the highest shape variation. This supports our interpretation of the previous result, with flakes from later in the reduction process showing greater shape variation as different types of cores were produced at the site.

@@

Looking at \@ref(fig:contribution-plots), the landmarks which contributed the most to each principal component were all located on the left side of our artefacts. This means that linking our initial flake measurements (maximum length, top width etc.) to either of the first two principal components is nearly impossible, as every one of our initial flake measurements (with the) one can see that there is no consistent pattern in which certain initial flake measurements  align closely with either the first or second principal components. Interestingly, the landmarks which contributed the most to each principal component were all located on the left side of our artefacts..

The lack of distinct morphological groupings and the increased variability of shape in the smaller, tertiary flakes is consistent with previous studies. Other studies which investigated the morphology of flake outlines from South East Asia concluded that there is no significant relationship between flake outline and flake use (Borel et al., 2013; Borel et al., 2017). <!-- summarise Borel et al. to show that we understand what they did and how they came to their conclusion, this might be better done in the introduction section  --> While our analysis does not rely upon outlines, both outlines and landmarks are methods for measuring the morphology of flaked stone tools, and as such, the same models which explain outline variability can be cautiously applied to our analysis. The high variability in flake shape, especially in the tertiary flakes, suggests that knappers were not looking to systematically create a small set of specific shapes of flake, with the amount of variability increasing as flakes are struck off of increasingly differed cores. This behavior is consistent with processes of expedient knapping, where a large number of flakes are created in a relatively free-form manner and useful flakes are picked up and used after knapping is finished (Holdaway & Douglass, 2012; McCall, 2012). Processes of expedient knapping suggest that stone tools were of secondary importance in the material culture of the knappers (Sillitoe & Hardy, 2003), relative to organic technologies. Mijares (2001) suggested stone tools were primarily used to create bamboo tools in South East Asia. A reliance upon bamboo tools, with stone flakes being an intermediary to produce bamboo tools, may explain the great variability in the lithic assemblages at Mau A.
